RTS AI Problems and Techniques

### 2014-12-09 ECGG_RTSProblems/ECGG_RTSProblems.tex
Add techniques section

This article aims to  provide an overview on what is  the state of the
art  in  RTS AI,  with  a  particular emphasis  on  the  work done  in
StarCraft.  It is organized as follows: 
{\color{blue} TODO}

@@

{\em StarCraft: Brood  War} is an immensely popular  RTS game released
in   1998  by   Blizzard  Entertainment.   StarCraft  is   set  in   a
science-fiction based universe where the player must choose one of the
three  races: Terran,  Protoss or  Zerg.  One of  the most  remarkable
aspects  of StarCraft  is  that  the three  races  are extremely  well
balanced:

\begin{itemize}
\item Terrans provide  units that are versatile and  flexible giving a
  balanced option between Protoss and Zergs.
\item  Protoss   units  have   lengthy  and   expensive  manufacturing
  processes, but they are strong and resistant.  These conditions make
  players follow a strategy of quality over quantity.
\item Zergs, the insectoid race, units are cheap and weak. They can be
  produced fast, encouraging players to overwhelm their opponents with
  sheer numbers.
\end{itemize}

% What about copyright problems with this screenshot?
% Figure~\ref{fig:StarCraft} shows  a screenshot of StarCraft  showing a
% player playing the Terran race. 
In order to win a StarCraft  game, players must first gather resources
(minerals  and Vespene  gas). As  resources become  available, players
need to allocate them for creating more buildings (which reinforce the
economy, and allow players to  create units or unlock stronger units),
research  new technologies  (in order  to  use new  unit abilities  or
improve the units)  and train attack units. Units  must be distributed
to  accomplish different  tasks  such as  reconnaissance, defense  and
attack.  While  performing all  of those tasks,  players also  need to
strategically understand the geometry of the  map at hand, in order to
decide where to place new buildings  (concentrate in a single area, or
expand  to  different  areas)  or where  to  set  defensive  outposts.
Finally, when  offensive units of  two players meet, each  player must
quickly maneuver each  of the units in order to  fight a battle, which
requires quick and reactive control of each of the units.

@@

A typical  StarCraft map is defined  as a rectangular grid,  where the
$width \times  height$ of  the map  is measured in  the number  of $32
\times 32$ squares of pixels, also  known as build tiles. However, the
resolution of  walkable areas is  in squares  of $8 \times  8$ pixels,
also known as  walk tiles. The typical dimensions for  maps range from
$64  \times 64$  to  $256 \times  256$ build  tiles.  Each player  can
control  up to  200 units  (plus  an unlimited  number of  buildings).
Moreover,  each different  race contains  between 30  to 35  different
types of units  and buildings, most of them with  a significant number
of  special abilities.  All these  factors together  make StarCraft  a
significant  challenge, in  which humans  are still  much better  than
computers.      For     instance,      in     the      game     ladder
iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked
by their current point totals ($E$ being the lowest possible rank, and
$A^+$  and  $Olympic$ being  the  second  highest and  highest  ranks,
respectively), the best  StarCraft AI bots are ranked  between $D$ and
$D^+$,  where average  amateur players  are ranked  between $C^+$  and
$B$. For comparison, StarCraft professional players are usually ranked
between $A^-$ and $A^+$.

From a theoretical point of view,  the state space of a StarCraft game
for a given map is enormous.  For example, consider a $128 \times 128$
map. At any given moment there might be between 50 to 400 units in the
map,  each of  which might  have a  complex internal  state (remaining
energy  and hit-points,  action  being executed,  etc.). This  quickly
leads to an immense number of  possible states (way beyond the size of
smaller games, such as Chess or Go). For example, just considering the
location of  each unit (with  $128 \times 128$ possible  positions per
unit),  and 400  units, gives  us  an initial  number of  $16384^{400}
\approx 10^{1685}$. If we add the  other factors playing a role in the
game, we obtain even larger numbers.

Another way to measure the complexity of the game is by looking at the
branching factor,  $b$, and the  depth of  the game, $d$,  as proposed
in~\cite{Gaby}, with  a total game  complexity of $b^d$. In  Chess, $b
\approx 35$  and $d \approx  80$. In more  complex games, like  Go, $b
\approx  30$ to  $300$, and  $d  \approx 150$  to $200$.  In order  to
determine the  branching factor in StarCraft  when an AI plays  it, we
must have in mind, that the  AI can issue actions simultaneously to as
many  units in  the  game as  desired. Thus,  considering  that, in  a
typical game, a player controls between 50 to 200 units, the branching
factor  would be  between $u^{50}$  and  $u^{200}$, where  $u$ is  the
average number of actions each unit can execute.  Estimating the value
of $u$ is not easy, since the  number of actions a unit can execute is
highly  dependent   on  the  context.   Let  us  make   the  following
assumptions: 1) at most 16 enemy units  will be in range of a friendly
unit (larger values  are possible, but unlikely), 2) when  an AI plays
StarCraft, it only makes sense to  consider movement in the 8 cardinal
directions per unit  (instead of assuming that the player  can issue a
``move'' command to anywhere in the map  at any point in time), 3) for
``build'' actions,  we consider that  SCVs (Terran worker  units) only
build in their  current location (otherwise, if they need  to move, we
consider  that  as  first  issuing  a  ``move''  action,  and  then  a
``build''), and  4) let's  consider only the  {\em Terran}  race. With
those assumptions,  units in  StarCraft can  execute between  1 (units
like ``Supply  Depots'', whose only  action is  to be ``idle'')  to 43
actions (Terran ``Ghosts''), with typical values around 20 to 30. Now,
if we have in mind that actions have cool-down times, and thus not all
units can  execute all of  the actions at every  frame, we can  take a
conservative estimation  of about $10$  possible actions per  unit per
game frame. This results in  a conservative estimate for the branching
factor  between $b  \in  [10^{50},10^{200}]$,  only considering  units
(ignoring the actions buildings can execute).
% Assuming  a conservative  value for  $u$  of about  $30$ (moving  to
% nearby positions, attacking units in range, using special abilities,
% etc.), we  obtain a  branching factor of  $b$ between  $30^{50}$ and
% $30^{200}$.
Now, to  compute $d$, we simply  consider the fact that  typical games
last for  about 25  minutes, which  results in  $d \approx  36000$ (25
minutes $\times$ 60 seconds $\times$ 24 frames per second).

>>>>

This article aims to  provide an overview on what is  the state of the
art  in  RTS AI,  with  a  particular emphasis  on  the  work done  in
StarCraft.  It is organized as follows: 
{\color{blue} TODO (or not)}

@@

{\color{blue}  Do we  need this  part?  Should  we talk  about RTS  AI
  problems  and   techniques  in  general,  but   take  examples  from
  StarCraft only, arguing this game is mainstream?}

{\em StarCraft: Brood  War} is an immensely popular  RTS game released
in   1998  by   Blizzard  Entertainment.   StarCraft  is   set  in   a
science-fiction based universe where the player must choose one of the
three  races: Terran,  Protoss or  Zerg.  One of  the most  remarkable
aspects  of StarCraft  is  that  the three  races  are extremely  well
balanced.

% \begin{itemize}
% \item Terrans provide  units that are versatile and  flexible giving a
%   balanced option between Protoss and Zergs.
% \item  Protoss   units  have   lengthy  and   expensive  manufacturing
%   processes, but they are strong and resistant.  These conditions make
%   players follow a strategy of quality over quantity.
% \item Zergs, the insectoid race, units are cheap and weak. They can be
%   produced fast, encouraging players to overwhelm their opponents with
%   sheer numbers.
% \end{itemize}

% What about copyright problems with this screenshot?
% Figure~\ref{fig:StarCraft} shows  a screenshot of StarCraft  showing a
% player playing the Terran race.

{\color{blue}  Thus,  merge the  following  paragraph  with the  first
  paragraph of the RTS section?}
In order to win a StarCraft  game, players must first gather resources
(minerals  and Vespene  gas). As  resources become  available, players
need to allocate them for creating more buildings (which reinforce the
economy, and allow players to  create units or unlock stronger units),
research  new technologies  (in order  to  use new  unit abilities  or
improve the units)  and train attack units. Units  must be distributed
to  accomplish different  tasks  such as  reconnaissance, defense  and
attack.  While  performing all  of those tasks,  players also  need to
strategically understand the geometry of the  map at hand, in order to
decide where to place new buildings  (concentrate in a single area, or
expand  to  different  areas)  or where  to  set  defensive  outposts.
Finally, when  offensive units of  two players meet, each  player must
quickly maneuver each  of the units in order to  fight a battle, which
requires quick and reactive control of each of the units.

@@

% A typical  StarCraft map is defined  as a rectangular grid,  where the
% $width \times  height$ of  the map  is measured in  the number  of $32
% \times 32$ squares of pixels, also  known as build tiles. However, the
% resolution of  walkable areas is  in squares  of $8 \times  8$ pixels,
% also known as  walk tiles. The typical dimensions for  maps range from
% $64  \times 64$  to  $256 \times  256$ build  tiles.  Each player  can
% control  up to  200 units  (plus  an unlimited  number of  buildings).
% Moreover,  each different  race contains  between 30  to 35  different
% types of units  and buildings, most of them with  a significant number
% of  special abilities.  All these  factors together  make StarCraft  a
% significant  challenge, in  which humans  are still  much better  than
% computers.      For     instance,      in     the      game     ladder
% iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked
% by their current point totals ($E$ being the lowest possible rank, and
% $A^+$  and  $Olympic$ being  the  second  highest and  highest  ranks,
% respectively), the best  StarCraft AI bots are ranked  between $D$ and
% $D^+$,  where average  amateur players  are ranked  between $C^+$  and
% $B$. For comparison, StarCraft professional players are usually ranked
% between $A^-$ and $A^+$.

From a theoretical point of view,  the state space of a StarCraft game
for a  given map is  enormous.  For  example, consider a  typical $128
\times 128$ map. At any given moment  there might be between 50 to 400
units in  the map, each of  which might have a  complex internal state
(remaining energy  and hit-points, action being  executed, etc.). This
quickly leads to an immense number  of possible states (way beyond the
size  of smaller  games,  such  as Chess  or  Go).  For example,  just
considering the location of each  unit (with $128 \times 128$ possible
positions per  unit), and  400 units,  gives us  an initial  number of
$16384^{400} \approx 10^{1685}$. If we add the other factors playing a
role in the game, we obtain even larger numbers.

% Another way to measure the complexity of the game is by looking at the
% branching factor,  $b$, and the  depth of  the game, $d$,  as proposed
% in~\cite{Gaby}, with  a total game  complexity of $b^d$. In  Chess, $b
% \approx 35$  and $d \approx  80$. In more  complex games, like  Go, $b
% \approx  30$ to  $300$, and  $d  \approx 150$  to $200$.  In order  to
% determine the  branching factor in StarCraft  when an AI plays  it, we
% must have in mind, that the  AI can issue actions simultaneously to as
% many  units in  the  game as  desired. Thus,  considering  that, in  a
% typical game, a player controls between 50 to 200 units, the branching
% factor  would be  between $u^{50}$  and  $u^{200}$, where  $u$ is  the
% average number of actions each unit can execute.  Estimating the value
% of $u$ is not easy, since the  number of actions a unit can execute is
% highly  dependent   on  the  context.   Let  us  make   the  following
% assumptions: 1) at most 16 enemy units  will be in range of a friendly
% unit (larger values  are possible, but unlikely), 2) when  an AI plays
% StarCraft, it only makes sense to  consider movement in the 8 cardinal
% directions per unit  (instead of assuming that the player  can issue a
% ``move'' command to anywhere in the map  at any point in time), 3) for
% ``build'' actions,  we consider that  SCVs (Terran worker  units) only
% build in their  current location (otherwise, if they need  to move, we
% consider  that  as  first  issuing  a  ``move''  action,  and  then  a
% ``build''), and  4) let's  consider only the  {\em Terran}  race. With
% those assumptions,  units in  StarCraft can  execute between  1 (units
% like ``Supply  Depots'', whose only  action is  to be ``idle'')  to 43
% actions (Terran ``Ghosts''), with typical values around 20 to 30. Now,
% if we have in mind that actions have cool-down times, and thus not all
% units can  execute all of  the actions at every  frame, we can  take a
% conservative estimation  of about $10$  possible actions per  unit per
% game frame. This results in  a conservative estimate for the branching
% factor  between $b  \in  [10^{50},10^{200}]$,  only considering  units
% (ignoring the actions buildings can execute).

% Assuming  a conservative  value for  $u$  of about  $30$ (moving  to
% nearby positions, attacking units in range, using special abilities,
% etc.), we  obtain a  branching factor of  $b$ between  $30^{50}$ and
% $30^{200}$.

% Now, to  compute $d$, we simply  consider the fact that  typical games
% last for  about 25  minutes, which  results in  $d \approx  36000$ (25
% minutes $\times$ 60 seconds $\times$ 24 frames per second).
